Off the beaten path: \\ Gaia reveals GD-1 stars perturbed off of the main stream track

### 2018-04-30 paper/gd1.tex
sec 3.2

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
We find stream ridge points by computing a running median in $\phi_2$ in steps
of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
second-order polynomial to the ridge points to find $\phi_{2,
\textrm{track}}(\phi_1)$.
We then define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model} (top panel) again shows the
high-confidence stream stars, and the two curved lines (blue) show the adopted
upper and lower boundary of the stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing median stream properties in overlapping windows:
we set the window size to $3^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each subsequent measurement.
Panels 3--5 in \figurename~\ref{fig:track-and-model} show the
background-subtracted stream surface density and median proper motion
components, $\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way (dark
lines), with the associated empirical scatters (shaded regions).
We estimate the local background surface density as an average of areas north
and south of the bounded stream region.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
Median stream properties computed from the proper motion and CMD filtered GD-1
stars, and the same for a toy model for the stream track.
\textit{Panel (a)}: Sky positions of candidate GD-1 stars selected using \gaia\
proper motions and \pans\ photometry.
Lines show a 2nd-order polynomial fit to the median stream track offset by $\pm
0.75^\circ$; the region between these lines is adopted as the nomial stream
region.
\textit{Panel (b)}: Sky positions of simulated star particles from a toy mock
stream model of the GD-1 stream track.
\textit{Panel (c)}: Background-subtracted surface-density estimates computed in
running, overlapping $3^\circ$ windows for both GD-1 data and toy model.
\textit{Panel (d)}: Median proper motion along the stream longitude computed in
the same bins as previous panel.
\textit{Panel (e)}: Same as Panel (d), but for proper motion in stream latitude.
}
\label{fig:track-and-model}
\end{figure}

@@

\subsection{Off-track features}
\label{sec:res_gap}

- overdensities apparent beyond the main stream track used in previous section -- if truly associated with the stream, this would be a novel insight into disruption from the \gaia\ mission
- since some contamination still remains present, it is possible that this is a variation in the MW population, just randomly coming into our selection boxes
- we thus perform an experiment and compare cmd and proper motions of spur and blob fields to stream and background fields to see which they are associated with

@@

- central panel of figure~\ref{fig:features} shows the map of GD-1, with fields of interest in shaded rectangles
- we first analyze the spur, region, stream same phi1, phi2 below, and an off-stream field in phi2
- for each of these fields, we show a pair of plots in the top row: cmd on the left and proper motions on the right.
- we first select on pm, and show cmd, and then select on cmd and show pm
- in both of these spaces, selection regions used to make the main catalog are shown in shaded orange.
- stream field (top left) has clear overdensity in the cmd selection box and clusters in proper motion
- on the other hand, the off field has some stars entering both selection boxes, but without any overdensities in either
- the spur field has fewer stars than the stream, but XX more than expected from the MW alone
- furthermore, the distribution in selection boxes is distributed similarly as in the stream field
- this confirms the association, but we note that in detail they differ
- median in pm offset from the stream field just below is XX, YY mas yr
- also dispersion different
- don't have enough stars to measure distance from the turn-off, so defer to future studies to measure by some other means, e.g., from photometry

- analogous paragraph for blob on the bottom

% - Spur vs. blob
% - Estimate velocity offset and significance
% - No indication of these features in previous work

% \figurename~\ref{fig:features} (middle panel) again shows the proper-motion and CMD selected stream stars, with interesting features

>>>>

To study the global properties of the stream as a function of stream longitude,
we extract stream stars around a low-order polynomial model for the median
stream latitude, $\phi_2$, as a function of stream longitude, $\phi_1$.
In detail, we find stream ridge points by computing a running median in $\phi_2$
in steps of $2^\circ$ with a window size of $4^\circ$ along $\phi_1$, then fit a
second-order polynomial to the ridge points to find $\phi_{2,
\textrm{track}}(\phi_1)$.
We define the stream as the region within $\left| \phi_2 - \phi_{2,
\textrm{track}}(\phi_1) \right| < 0.75^\circ$.
\figurename~\ref{fig:track-and-model} (top panel) again shows the
high-confidence stream stars, and the two curved lines (blue) show the adopted
upper and lower boundary of the stream region.

We extract properties of the stream from the region defined above as a function
of $\phi_1$ by computing stream properties in overlapping windows:
we set the window size to $4^\circ$ in $\phi_1$, and shift the window center by
$1^\circ$ in longitude for each measurement.
Rows 2--5 in \figurename~\ref{fig:track-and-model} show the background-subtracted
stream surface density, \todo{$\Sigma$?}, the median latitude, $\phi_2$, and median proper motions,
$\mu_{\phi_1}$ and $\mu_{\phi_2}$, extracted in this way, and the associated empirical scatters
(blue lines and shaded regions).
We estimate the local background surface density as an average of areas north and south from the bounded stream region.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{track_observables.pdf}
\end{center}
\caption{%
\todo{APW}
}
\label{fig:track-and-model}
\end{figure}

@@

\subsection{Off-track features}
\label{sec:res_gap}
In this section we test whether the overdensities apparent beyond the main stream track (Figure~\ref{fig:selection}) are truly associated with the GD-1 stream.
% this would be a novel insight into disruption from the \gaia\ mission
Some contamination still remains present in our selection, so it is possible that these features are due to the Milky Way population randomly over-populating our selection boxes.
To assess the origin of these features, we compare color-magnitude diagrams and proper motions of the spur and blob fields to the stream and background fields.

@@

The central panel of Figure~\ref{fig:features} shows the map of GD-1, with fields of interest marked in shades of purple for spur-related fields, and shades of blue for blob-related fields.
In both cases, we analyze the stream region (darkest shade), overdensity off from the stream (medium shade) and use a field on the opposite side of the stream from the overdensity as a control (lightest shade).
The $\phi_1$ range is the same for all three fields associated with a given feature, and the control and feature field have the same area, while the stream field is smaller to avoid cross-contamination.

For each of the examined fields, we present the color-magnitude diagram of stars selected on proper motions (left) and proper motions of photometrically selected stars (right), with spur fields on the top and blob fields on the bottom of Figure~\ref{fig:features}.
In both rows, arranged from left to right are the stream, the feature, and the control field.
Stars selected from the stream fields show a clear overdensity in the CMD selection box and they cluster in proper motions.
In detail, the two stream fields have a clear relative offset in proper motions, as expected from the velocity gradients shown in Figure~\ref{fig:track-and-model}.
Furthermore, the main-sequence turn-off of the $\phi_1\sim-15^\circ$ field appears to be fainter than the $\phi_1\sim-35^\circ$, consistent with the distance gradient across the stream measured by \citet{Koposov:2010}.
On the other hand, the control fields have some stars entering both the CMD and proper-motion selection boxes, but without any significant overdensities in either.
CMDs of the spur and blob fields have more stars in the selection box than their respective control fields, and the distribution within the box is qualitatively similar to that of the corresponding stream field, suggesting a similar distance.
In proper motions, the spur and blob fields also display properties similar to the adjacent stream fields, with clear overdensities of stars in the selection box.

Stars from the spur and the blob are similar to those along the main stream track in the color-magnitude diagram and in proper motions, which confirms their association to GD-1.
The proximity of these off-stream features to large gaps in the stream is suggestive of the gaps being the origin site, further implied by hints of relative velocity differences; for example the median proper motion of the blob is offset from the adjacent stream field by $\approx0.3$\,mas\,yr$^{-1}$ in the positive $\mu_{\phi_1}$ direction.
We defer detailed modeling of these features to infer their co-evolution with the stream for a future study.
